---
title: "Data Wrangling"
author: "ECON145 Instructor"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    theme: readable
    self_contained: TRUE
    toc: TRUE
    number_sections: TRUE
    df_print: paged
---

"dplyr" and "tidyr" packages are powerful R packages for data manipulation.
```{r}
library(dplyr)
library(tidyr)
library(ggplot2)
```

Core tidyverse Packages:
ggplot2: For data visualization, creating graphics and plots based on the grammar of graphics.
dplyr: For data manipulation, including filtering, selecting, mutating, summarizing, and arranging data.
tidyr: For tidying data, making datasets easy to manipulate (e.g., reshaping data, handling missing values).

```{r}
getwd()
```


```{r}
data <- read.csv("2023_WorldBank.csv", stringsAsFactors = FALSE)
head(data)
str(data)
colnames(data)
summary(data)
```

"Country.Code", "Indicator.Name", "Indicator.Code" are not important. We drop them with dplyr select statement. 
You can also use baseR alternative: `data <- data[, !names(data) %in% c("Country.Code", "Indicator.Name", "Indicator.Code")]`

Data appears to be in wide format, each year is in a column. We can use tidyr to make this into a tidy data and recode column names as the year of observation. 

tidyr is another R package from the tidyverse family, designed to help users clean and organize data into a tidy format. The primary goal of tidyr is to make data "tidy," meaning that each variable is a column, each observation is a row, and each value is a single cell. tidyr works closely with other packages like dplyr to simplify data manipulation workflows.

*Key functions of tidyr include:*
pivot_longer(): Converts wide data (where columns are variables) into long format (where each row represents a single observation of a variable).
pivot_wider(): The reverse of pivot_longer(), it converts long data into wide format.
separate(): Splits a single column into multiple columns based on a delimiter.
unite(): Combines multiple columns into a single column.

```{r}
# remove unneccesary columns
data <- dplyr::select(data, -c(Country.Code, Indicator.Name, Indicator.Code))

# Pivot to tidy format where each Column is a Variable and each Row is an Observation (Life Expentancy)
data_long <- pivot_longer(data, 
                          cols = X1960:X2021, 
                          names_to = "Year", 
                          values_to = "Life.Exp")

# The 'Year' column will have year names as characters starting with X. We remove the "X" and convert them to integers
data_long$Year <- as.integer(sub("X", "", data_long$Year))

str(data_long)
colnames(data_long)
```
dplyr is an R package used for data manipulation and transformation. It is part of the tidyverse, a collection of R packages designed for data science. dplyr provides a set of functions that make it easier to work with data frames and manipulate them in a clear, concise, and efficient way.

Key features of dplyr include:
Pipes (%>%): Allows chaining multiple operations in a readable manner.

Core Functions:
select(): Choose columns.
filter(): Filter rows based on conditions.
mutate(): Create new columns or modify existing ones.
summarize(): Aggregate data, often combined with group_by() to create group-wise summaries.
group_by(): Group data by one or more variables, which is useful for applying functions like summarize().

```{r}
# applies the sum function for all NA values to all columns in data_long. check for NA values. Life Expectancy has 626 NA's
data_long %>%
  summarise_all(~ sum(is.na(.)))

# Seems like all the missing values appear in Life.Exp. We check the unique values for Country.Name and Year first
unique(data_long$Country.Name)
unique(data_long$Year)

# Since all the values look correct, we check the missing value across each Country for Life.Exp column
temp_df <-data_long %>%
  group_by(Country.Name) %>%
  summarise(Missing_Life_Exp = sum(is.na(Life.Exp)))
temp_df

# To show only the rows with non-zero Missing Counts
data_long %>%
  group_by(Country.Name) %>%
  summarise(Missing_Life_Exp = sum(is.na(Life.Exp))) %>%
  filter(Missing_Life_Exp > 0)

# Here we remove all rows with missing Life.Exp as they are not important for our analysis. na.omit is using baseR.
data_long <- na.omit(data_long)
length(unique(data_long$Country.Name))
length(unique(data_long$Year))
data_long

```

Next we check the summary statistics for the data across all countries. We again use `group_by` and `summarize` functions in `dplyr`. We can then use the summary statistics table to check with domain experts on the validity of the data.

```{r}
country_summary_stats <- data_long %>%
  group_by(Country.Name) %>%
  summarize(
    Mean = mean(Life.Exp, na.rm = TRUE),
    Median = median(Life.Exp, na.rm = TRUE),
    SD = sd(Life.Exp, na.rm = TRUE),
    Min = min(Life.Exp, na.rm = TRUE),
    Max = max(Life.Exp, na.rm = TRUE),
    Count = n(),
    Missing = sum(is.na(Life.Exp))
  )

country_summary_stats
length(unique(data_long$Country.Name))
```

There should be 62 years of data for each country. We remove countries with less than 62 years of data.
The pull() function extracts a specific column from the filtered data.
Here, it pulls the values of the Country.Name column from the filtered rows where Count is less than 62.

```{r}
# 1. Filter out country names with Count < 62
countries_to_remove <- country_summary_stats %>%
  filter(Count < 62) %>%
  pull(Country.Name)

countries_to_remove

# 2. Filter out these countries from data_long
data_long_filtered <- data_long %>%
  filter(!Country.Name %in% countries_to_remove)
length(unique(data_long_filtered$Country.Name))
```

We can also plot some charts using ggplot to visualize the data and use it to discuss with domain experts.
We can plot with ggplot:
1. Histogram of Life.Exp for all the data
2. Boxplot of Life.Exp by Year
3. Boxplot of Life.Exp by Country for selected Countries
4. Time series plots for selected countries

```{r}
# Histogram
ggplot(data_long_filtered, aes(x = Life.Exp)) +
  geom_histogram(binwidth = 1, fill = "blue", color = "black") +
  theme_minimal() +
  labs(title = "Histogram of Life Expectancy", x = "Life Expectancy", y = "Count")
```

```{r}
# Boxplot of Life.Exp by Year
ggplot(data_long_filtered, aes(x = as.factor(Year), y = Life.Exp)) +
  geom_boxplot() +
  theme(axis.text.x = element_text(angle = 90)) +
  labs(title = "Life Expectancy by Year", x = "Year", y = "Life Expectancy")

```

```{r}
# Boxplot of Life.Exp by Country for top 20 highest Mean Life Expectancy across the years
# 1. Extract the Top 20 countries' Country.Name with `dplyr`
top_20 <- data_long_filtered %>%
  group_by(Country.Name) %>%
  summarize(MeanLifeExp = mean(Life.Exp, na.rm = TRUE)) %>%
  arrange(desc(MeanLifeExp)) %>%
  top_n(20, MeanLifeExp) %>%
  pull(Country.Name)
top_20
```

```{r}
# 2. Filter these countries with `dplyr`
data_top_20 <- data_long_filtered%>%
  filter(Country.Name %in% top_20)
unique(data_top_20$Country.Name)
# List excludes Singapore

# Include Singapore in the list for comparison
data_top_20_with_sg <- data_long_filtered %>%
  filter(Country.Name %in% top_20 | Country.Name == 'Singapore')

data_top_20_with_sg
```

```{r}
# 3. Plot the Boxplot for Top 20 countries and Singapore
ggplot(data_top_20_with_sg, aes(x = Country.Name, y = Life.Exp)) +
  geom_boxplot() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  labs(title = "Top 20 Countries by Mean Life Expectancy", x = "Country", y = "Life Expectancy")
```

```{r}
# Reorder by descending mean Life.Exp (`-Life,Exp` for descending)
ggplot(data_top_20_with_sg, aes(x = reorder(Country.Name, -Life.Exp, FUN = mean), y = Life.Exp)) +
  geom_boxplot() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  labs(title = "Top 20 Countries by Mean Life Expectancy", x = "Country", y = "Life Expectancy")
```
We can also plot time series for the Top 20 countries to visualize the trends.

```{r}
# Create the time series plot
tseries <-ggplot(data_top_20, aes(x = Year, y = Life.Exp, color = Country.Name)) +
  geom_line() +
  theme_minimal() +
  labs(title = "Life Expectancy Over Time",
       x = "Year",
       y = "Life Expectancy",
       color = "Country") 
tseries

```

```{r}
# Add Singapore in the comparison
tseries + 
  geom_line(data = filter(data_top_20_with_sg, Country.Name == "Singapore"), 
            aes(x = Year, y = Life.Exp), 
            color = "black", 
            size = 1.5)
```

Congratulations! We have completed a quick introduction to dplyr, tidyr and ggplot. Remember, the best way **to learn** is to get your hands dirty, google for sample codes, read R documentation and learn when you are trying to solve problems.