---
title: "Univariate Association"
author: "ECON145 Instructor"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    theme: readable
    self_contained: TRUE
    toc: TRUE
    number_sections: TRUE
    df_print: paged
---

```{r}
library(ggplot2)
```


Refer to the Framingham Heart Study micro-dataset

```{r}
framingham <- read.csv("Framingham.csv")
str(framingham)
```

Here we categorize the patients into elderly (65 years old and above), pre-elderly (50 to less than 65 years old) and non-elderly (less than 50 years old)

```{r}
framingham$age_category <- cut(framingham$age,
                          breaks = c(-Inf, 50, 65, Inf),
                          labels = c("Less than 50", "50 to less than 65", "65 and above"))
summary(framingham)
```

Suppose want to study if there is any statistically significant differences across elderly, pre-elderly and non-elderly groups for sysBP (Systolic BP). Elevated systolic blood pressure is a risk factor for heart disease and stroke.)

```{r}
ggplot(framingham, aes(x = age_category, y = sysBP)) +
  geom_boxplot() +
  labs(title = "Boxplot of Systolic BP across Age Categories",
       x = "Age Category",
       y = "Systolic BP") +
  theme_minimal() +
  theme(legend.title = element_blank())
summary(framingham$sysBP)
```

The Median values seemed higher for pre-elderly and elderly age groups. We can perform one-way ANOVA to check if there are any statistically significant differences. The ANOVA test assumes that the data follows a normal distribution and that the variances are equal across groups (homogeneity of variances).

```{r}
anova_result <- aov(sysBP ~ age_category, data = framingham)
summary(anova_result)
?aov

```

We should also test the IID Normal assumptions.
(1) Normality Assumption

```{r}
# Shapiro-Wilk normality test for each group
by(framingham$sysBP, framingham$age_category, shapiro.test)

# Q-Q plots for each group
ggplot(framingham, aes(sample = sysBP)) +
  facet_wrap(~age_category) +
  geom_qq() +
  geom_qq_line()
```

(2) Test for Homoscedasticity

```{r}
# Levene's test for homogeneity of variances
library(car)
leveneTest(sysBP ~ age_category, data = framingham)
```

Assumptions seems to be violated. Try non-parametric alternative.

```{r}
# Kruskal-Wallis test as a non-parametric alternative to ANOVA
kruskal.test(sysBP ~ age_category, data = framingham)
```

Conclusion is still the same as for ANOVA