---
title: "Framingham Heart Study Using Linear Regression"
author: "ECON145 Instructor"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    theme: readable
    self_contained: TRUE
    toc: TRUE
    number_sections: TRUE
    df_print: paged
---



Refer back to the Framingham Heart Study dataset. Load libraries.

```{r}
library(ggplot2)
framingham <- read.csv("Framingham.csv")
str(framingham)
```

We look at Gender association with the response of "TenYearCHD" without concerning ourselves with the other covariates in a univariate analysis. `framingham$male` captures the gender with male = `1` and female = `0`. We first convert this binary variable into a factor data type.

```{r}
unique(framingham$male)
framingham$male = as.factor(framingham$male)
framingham_small <- framingham[c("male","TenYearCHD")]
str(framingham_small)

```

Visualize the data with bar chart.

```{r}
ggplot(data = framingham_small, aes(x = male, fill = factor(TenYearCHD))) +
  geom_bar(position = "dodge") +
  labs(x = "Gender", y = "Count", fill = "TenYearCHD") +
  scale_fill_manual(values = c("0" = "blue", "1" = "red"), labels = c("No CHD", "CHD")) +
  theme_minimal() +
  ggtitle("Counts of CHD by Gender")
```

Visualize the data with stacked bar chart.

```{r}
ggplot(data = framingham_small, aes(x = male, fill = factor(TenYearCHD))) +
  geom_bar(position = "fill") +
  labs(x = "Gender", y = "Proportion", fill = "TenYearCHD") +
  scale_y_continuous(labels = scales::percent) +
  scale_fill_manual(values = c("0" = "blue", "1" = "red"), labels = c("No CHD", "CHD")) +
  theme_minimal() +
  ggtitle("Proportion of CHD Occurrence by Gender")
```

You can also try to visualize with a mosaic plot with the `vcd` package. Mosaic plot is like your contingency table.

```{r}
library(vcd)
# Mosaic plot for visualizing the association between gender and CHD
mosaic(~ male + TenYearCHD, data = framingham_small,
       main = "Mosaic Plot of CHD Occurrence by Gender",
       xlab = "Gender",
       ylab = "TenYearCHD",
       color = TRUE)

```

Summarize the data into a contingency table then call the `chi.sq` test.
```{r}
framingham_small$gender <- factor(framingham_small$male, levels = c(0, 1), labels = c("female", "male"))
framingham_small$TenYearCHD <- factor(framingham_small$TenYearCHD, levels = c(0, 1), labels = c("NoCHD", "CHD"))

# Chi-squared codes
tableCHD <- table(framingham_small$gender, framingham_small$TenYearCHD)
tableCHD
chisq.test(tableCHD)
```

Congratulations! You have successfully learnt how to conduct a simple chi-squared test for categorical variables.

